---
title: "Tanmay Sharma - Data Science Portfolio"
---

# Welcome 👋

Hi, I’m **Tanmay Sharma**, a Master’s student at [Baruch College](https://zicklin.baruch.cuny.cuny.edu/).  
I am passionate about turning data into actionable insights. My work spans data analysis, visualization, and predictive modeling, helping uncover patterns and drive smarter decisions. I enjoy exploring real-world datasets, building interactive tools, and communicating findings in a clear, impactful way.

You can also connect with me on [LinkedIn](https://www.linkedin.com/in/tanmay-sharma-3146bb160).

---

## Interactive Map

```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
  options(repos=c(CRAN="https://cloud.r-project.org"))
  install.packages("leaflet")
  stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- 40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```

---

## Want to test your data science knowledge? Try this quiz!

---
title: "Data Science Quiz"
format:
  learnr::tutorial:    # <-- enables interactive quiz
    progressive: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
```

## Mini Data Science Quiz

### Question 1
```{r}
quiz(
  question("Which of these is **NOT** a supervised learning algorithm?",
           answer("Linear regression"),
           answer("K-means clustering", correct = TRUE),
           answer("Random forest"),
           answer("Logistic regression"),
           allow_retry = TRUE)
)
```

### Question 2
```{r}
quiz(
  question("What does the R function `set.seed()` do?",
           answer("It improves model accuracy"),
           answer("It ensures reproducible random results", correct = TRUE),
           answer("It normalizes the data"),
           allow_retry = TRUE)
)
```

### Question 3
```{r}
quiz(
  question("Which of the following best describes **tidy data**?",
           answer("Each column is a variable and each row is an observation", correct = TRUE),
           answer("All numeric values are standardized"),
           answer("Data stored in a single CSV file"),
           allow_retry = TRUE)
)
```
