---
title: Mini Project 01 - Netflix Analysis
format: html
---
**Step 1:** Acquiring Data From Netflix's TuDum Top 10. Click the *Code* Icon to See Full Code Throughout This Mini-Project

```{r, echo=TRUE}
#| code-fold: true
if(!dir.exists(file.path("data", "mp01"))){
    dir.create(file.path("data", "mp01"), showWarnings=FALSE, recursive=TRUE)
}

GLOBAL_TOP_10_FILENAME <- file.path("data", "mp01", "global_top10_alltime.csv")

if(!file.exists(GLOBAL_TOP_10_FILENAME)){
    download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv", 
                  destfile=GLOBAL_TOP_10_FILENAME)
}

COUNTRY_TOP_10_FILENAME <- file.path("data", "mp01", "country_top10_alltime.csv")

if(!file.exists(COUNTRY_TOP_10_FILENAME)){
    download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv", 
                  destfile=COUNTRY_TOP_10_FILENAME)
}
```

**Step 2:** Data Import & Preparation

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
if(!require("tidyverse")) install.packages("tidyverse")
library(readr)
library(dplyr)

GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME)
```

**Step 3:** Data Cleaning

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
GLOBAL_TOP_10 <- GLOBAL_TOP_10 |>
  mutate(season_title = if_else(season_title == "N/A", NA_character_, season_title))

COUNTRY_TOP_10 <- read_tsv(
  COUNTRY_TOP_10_FILENAME,
  na = "N/A"
)
```

**Step 4:** Initial Data Exploration

```{r, echo=TRUE}
#| code-fold: true
library(DT)
GLOBAL_TOP_10 |> 
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE))
```

**Formatting Column Names**. Also setting up a *clean* table for Question #5 as well as for interactive tables used subsequently. Click the *Code* Icon Below to See Full Code

```{r, echo = TRUE}
#| code-fold: true
library(stringr)
format_titles <- function(df){
    colnames(df) <- str_replace_all(colnames(df), "_", " ") |> str_to_title()
    df
}

# THIS IS THE NEW CHUNK; Clean country table column names for Question 5 and interactive tables
COUNTRY_TOP_10_CLEAN <- COUNTRY_TOP_10 %>% 
  format_titles()

colnames(COUNTRY_TOP_10_CLEAN)[1] <- "Country"
colnames(COUNTRY_TOP_10_CLEAN)[colnames(COUNTRY_TOP_10_CLEAN) == "Show Title"] <- "Show_Title"
colnames(COUNTRY_TOP_10_CLEAN)[colnames(COUNTRY_TOP_10_CLEAN) == "Cumulative Weeks In Top 10"] <- "Cumulative_weeks_In_Top_10"
#NEW CHUNK ENDS HERE

GLOBAL_TOP_10 |> 
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

Dropping **season_title** to get closer to publication quality titles

```{r, echo = TRUE}
#| code-fold: true
GLOBAL_TOP_10 |> 
    select(-season_title) |>
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

Converting runtime from *hours* to *minutes*:

```{r, echo=TRUE}
#| code-fold: true
GLOBAL_TOP_10 |>
  mutate(`runtime_(minutes)` = round(60 * runtime)) |>
  select(-season_title,
         -runtime) |>
  format_titles() |>
  head(n=20) |>
  datatable(options=list(searching=FALSE, info=FALSE)) |>
  formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

**Now Beginning Exploratory Data Analysis Section.** The code fold below shows all code for Questions 1-10 given on Prof. Weylandt's website. Markdown is subsequently used for inline text and to "call" variables out of the code.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
library(dplyr)

#Setting variable num_countries for Question 1
num_countries <- COUNTRY_TOP_10 %>%
  distinct(country_name) %>%
  nrow()

#Non english top film for Question 2
non_english_top <- GLOBAL_TOP_10 %>%
  mutate(Language = str_extract(category, "\\((.*)\\)")) %>%
  mutate(Language = str_replace_all(Language, "[()]", "")) %>%
  filter(str_detect(category, "Film"), Language != "English") %>%
  group_by(show_title) %>%
  summarise(total_weeks = sum(cumulative_weeks_in_top_10, na.rm = TRUE)) %>%
  arrange(desc(total_weeks)) %>%
  slice(1)

#Longest film and runtime for Question 3
longest_film <- GLOBAL_TOP_10 %>%
  filter(str_detect(category, "Film")) %>%
  arrange(desc(runtime)) %>%
  slice(1)

library(scales)

#Program (per category) with most total hours of global viewership for Question 4
top_hours <-GLOBAL_TOP_10 %>%
  group_by(category, show_title) %>%
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE), .groups = "drop") %>%
  group_by(category) %>%
  slice_max(total_hours, n=1) %>%
  ungroup() %>%
  format_titles() %>% #this converts column names to title case
  mutate(`Total Hours` = comma(`Total Hours`))

#Creating variable for longest run in top 10 for country for Question 5
longest_country_run <- COUNTRY_TOP_10_CLEAN %>% 
  filter(str_detect(Category, "TV")) %>% #match both English & Non-English TV
  group_by(Show_Title, Country) %>%
  summarise(run_weeks = sum(Cumulative_weeks_In_Top_10, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(run_weeks)) %>%
  slice(1)

#Creating variable for Country with < 200 weeks of history for Question 6
country_history <- COUNTRY_TOP_10_CLEAN %>%
  group_by(Country) %>%
  summarise(
    total_weeks = n_distinct(Week),
    last_week = max(as.Date(Week), na.rm = TRUE)
  ) %>%
  filter(total_weeks < 200)

#Creating variable for total viewership of Squid Game for Question 7
squid_game_hours <- GLOBAL_TOP_10 %>%
  filter(str_detect(show_title, "Squid Game")) %>%
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE))

  #Converting this variable to be read easily
squid_game_total <- formatC(squid_game_hours$total_hours, format = 'f', big.mark = ",", digits = 0)

#Creating variable for Red Notice Views for Question 8

library(lubridate)

red_notice_views <- GLOBAL_TOP_10 %>%
  filter(str_detect(show_title, "Red Notice"), year(week) == 2021) %>%
  mutate(approx_views = weekly_hours_viewed / (1+58/60)) %>%
  summarise(total_views = sum(approx_views, na.rm = TRUE))

  #Converting to readable manner
red_notice_total <- formatC(red_notice_views$total_views, format = "f", big.mark = ",", digits = 0)

#Films that reached #1 in US after debuting lower for Question 9
us_films <- COUNTRY_TOP_10_CLEAN %>%
  filter(Country == "United States", str_detect(Category, "Film"))

# Getting debut rank for each film and whether it later hit #1
film_climbers <- us_films %>%
  group_by(Show_Title) %>%
  summarise(
    debut_week = min(Week, na.rm = TRUE),
    debut_rank = `Weekly Rank`[which.min(Week)],
    ever_number1 = any(`Weekly Rank` == 1, na.rm = TRUE),
    last_number1_week = if (any(`Weekly Rank` == 1, na.rm = TRUE)) {
      max(Week[`Weekly Rank` == 1], na.rm = TRUE)
    } else {
      as.Date(NA)
    },
    .groups = "drop"
  ) %>%
  filter(debut_rank != 1 & ever_number1 == TRUE)

#For Question 10; TV show/season with widest debut across countries
tv_debut <- COUNTRY_TOP_10_CLEAN |>
  filter(str_detect(Category, "TV")) |>
  group_by(Show_Title) |>
  summarise(debut_week = min(Week, na.rm = TRUE), .groups = "drop")

tv_debut_counts <- COUNTRY_TOP_10_CLEAN |>
  inner_join(tv_debut, by = c("Show_Title", "Week" = "debut_week")) |>
  group_by(Show_Title, Week) |>
  summarise(num_countries = n_distinct(Country), .groups = "drop") |>
  arrange(desc(num_countries)) |>
  slice(1)
```

#### 1. How many countries does Netflix operate in?

Netflix operates in `r num_countries` countries.

#### 2. Which non-English-language film has spent the most cumulative weeks in the global top 10?

The non-English-language film with the most cumulative weeks in global Top 10 is `r non_english_top$show_title` with `r non_english_top$total_weeks` weeks.

#### 3. What is the longest film (English or non-English) to have ever appeared in the Netflix global Top 10? How long is it in minutes?

The longest film in the Global Top 10 is `r longest_film$show_title` with a runtime of `r round(longest_film$runtime*60)` minutes.

#### 4. For each of the four categories, what program has the most total hours of global viewership?

Programs with the most total hours per category:

`r DT::datatable(top_hours, options = list(searching=FALSE, info=FALSE))`

#### 5. Which TV show had the longest run in a country's Top 10? How long was this run and in what country did it occur?

The TV show with the longest run in a country is `r longest_country_run$Show_Title` in `r longest_country_run$Country` with `r longest_country_run$run_weeks` weeks.

#### 6. Netflix provides over 200 weeks of service history for all but one country in our data set. Which country is this and when did Netflix cease operations in that country?

The country with fewer than 200 weeks of history is
`r country_history$Country`, with last recorded week on `r country_history$last_week`.

#### 7. What is the total viewership of the TV show Squid Game? Note that there are three seasons total and we are looking for the total number of hours watched across all seasons.

Total viewership of Squid Game across all seasons is `r squid_game_total` hours.

#### 8. Approximate number of views of Red Notice in 2021 is
`r red_notice_total` views

#### 9. How many Films reached Number 1 in the US but did not originally debut there? That is, find films that first appeared on the Top 10 chart at, e.g., Number 4 but then became more popular and eventually hit Number 1? What is the most recent film to pull this off?

A total of `r nrow(film_climbers)` films reached **#1 in the US** after debuting at a lower rank. The most recent film to do this was `r film_climbers$Show_Title[which.max(film_climbers$last_number1_week)]`,
which hit #1 in the week of `r max(film_climbers$last_number1_week, na.rm = TRUE)`.

#### 10. Which TV show/season hit the top 10 in the most countries in its debut week? In how many countries did it chart?

The TV show/season that hit the Top 10 in the most countries during its debut week is
`r tv_debut_counts$Show_Title`, which charted in `r tv_debut_counts$num_countries` countries.

## Press Release 1 : Stranger Things Season 5

**Season 5 of Stranger Things Obliterates Records as Global Audiences Return back to Hawkins**

Netflix is pleased to announce the return of Stranger Things with its' much anticipated Season 5, establishing the show as one of the platform's most notable franchises. In the week that it debuted, the show charted in **over 90 countries**, claiming the #1 spot in overall categories as well as **English TV**. Through its run to the top, *Stranger Things* has amassed to over **billions of hours viewed**, with its latest season garnering over **300 million hours** in the first month following release. The show holds many different records including the **widest TV debut across countries**, a notable feat showcasing its global cultural impact. As the town of Hawkins enters its last chapter, fans across the globe continue to demonstrate that Stranger Things is more than just a show - it's a phenomenon.

```{r, warning=FALSE, message=FALSE, echo=TRUE}
#| code-fold: true
library(dplyr)
library(ggplot2)
library(stringr)

# Step 1: Filter TV shows from COUNTRY_TOP_10_CLEAN
tv_country <- COUNTRY_TOP_10_CLEAN %>%
  filter(Category == "TV", !is.na(Week))  # only valid weeks

# Step 2: Get category/language from GLOBAL_TOP_10 safely
global_tv_cat <- GLOBAL_TOP_10 %>%
  filter(str_detect(category, "TV")) %>%      # TV only
  select(show_title, category) %>%
  distinct(show_title, category)             # avoid many-to-many join

# Step 3: Join language info
tv_country <- tv_country %>%
  left_join(global_tv_cat,
            by = c("Show_Title" = "show_title"),
            relationship = "many-to-many")

# Step 4: Keep only English TV
tv_country_eng <- tv_country %>%
  filter(category == "TV (English)")

# Step 5: Compute debut week per show
tv_debut_english <- tv_country_eng %>%
  group_by(Show_Title) %>%
  summarise(debut_week = min(Week, na.rm = TRUE), .groups = "drop")

# Step 6: Count number of countries per show at debut
tv_debut_counts_eng <- tv_country_eng %>%
  inner_join(tv_debut_english, by = c("Show_Title", "Week" = "debut_week")) %>%
  group_by(Show_Title, Week) %>%
  summarise(num_countries = n_distinct(Country), .groups = "drop") %>%
  arrange(desc(num_countries)) %>%
  slice_head(n = 10)  # top 10

# Step 7: Highlight Stranger Things
tv_debut_counts_eng$highlight <- ifelse(tv_debut_counts_eng$Show_Title == "Stranger Things",
                                        "Stranger Things", "Other")

# Step 8: Plot
ggplot(tv_debut_counts_eng, aes(x = reorder(Show_Title, num_countries),
                               y = num_countries, fill = highlight)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Stranger Things" = "red", "Other" = "grey70")) +
  labs(
    title = "Top 10 Widest TV Debuts on Netflix (English TV)",
    subtitle = "Number of countries appearing in Top 10 during debut week",
    x = "TV Show",
    y = "Countries in Debut Week",
    fill = "Show"
  ) +
  theme_minimal(base_size = 12)
```

## Press Release 2: Netflix’s Commercial Success in India

**Through the Introduction of Hindi-Language Programs, Netflix Experiences Rapid Growth in India**

Through investing in region-specific content, Netflix sees huge growth in viewership in its platform. Specifically, both Hindi-language TV shows and films have achieved astounding popularity throughout India. These hits are helping Netflix in reaching new audiences and are also solidifying Netflix's position in one of the world's biggest media markets. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
library(dplyr)
library(ggplot2)
library(stringr)
library(ggtext)  # for element_markdown()

# Step 1: Filter Hindi/non-English programs in India
india_non_english <- COUNTRY_TOP_10_CLEAN %>%
  filter(Country == "India", !is.na(Week)) %>%
  left_join(
    GLOBAL_TOP_10 %>% select(show_title, category) %>% distinct(),
    by = c("Show_Title" = "show_title")
  ) %>%
  filter(str_detect(category, "Non-English")) %>%
  group_by(Show_Title) %>%
  summarise(total_weeks = sum(Cumulative_weeks_In_Top_10, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_weeks))

# Step 2: Select top Hindi programs
hindi_programs <- c("RRR (Hindi)", "The Great Indian Kapil Show", 
                    "Sooryavanshi", "Haseen Dillruba", "Khakee: The Bihar Chapter")

top_hindi <- india_non_english %>%
  filter(Show_Title %in% hindi_programs)

# Step 3: Highlight the top program
top_hindi$highlight <- ifelse(top_hindi$total_weeks == max(top_hindi$total_weeks),
                              "Top Program", "Other")
top_hindi$highlight <- factor(top_hindi$highlight, levels = c("Top Program", "Other"))

# Step 4: Plot top Hindi programs in India with bold y-axis labels and colored top program
top_program <- top_hindi$Show_Title[top_hindi$highlight=="Top Program"]

ggplot(top_hindi, aes(x = reorder(Show_Title, total_weeks),
                       y = total_weeks, fill = highlight)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Top Program" = "red", "Other" = "grey70")) +
  labs(
    title = "Top Hindi-Language Programs on Netflix India",
    subtitle = "Cumulative Weeks in Top 10",
    x = "Program",
    y = "Cumulative Weeks in Top 10",
    fill = "Highlight"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_markdown()
  ) +
  scale_x_discrete(labels = function(x) {
    ifelse(x == top_program,
           paste0("<span style='color:red;'><b>", x, "</b></span>"),
           paste0("<b>", x, "</b>"))
  })

# Step 5: Estimate audience
total_weeks_top5 <- sum(top_hindi$total_weeks)
average_weeks_per_user <- 4   # assumption: 1 program/week per user
estimated_customers_india <- total_weeks_top5 / average_weeks_per_user
estimated_customers_friendly <- format(round(estimated_customers_india, 0), big.mark = ",")

# Compute statistics in the code chunk
total_weeks_top5 <- sum(top_hindi$total_weeks)
#new section here
khakee <- top_hindi$total_weeks[top_hindi$Show_Title == "Khakee: The Bihar Chapter"]
khakee_weeks <- top_hindi$total_weeks[top_hindi$Show_Title == "Khakee: The Bihar Chapter"]
khakee_weeks_friendly <- format(khakee_weeks, big.mark = ",")
#end of new section - remove if necessary
average_weeks_per_user <- 4   # assumption: 1 program/week per user
estimated_customers_india <- total_weeks_top5 / average_weeks_per_user
estimated_customers_friendly <- format(round(estimated_customers_india, 0), big.mark = ",")
total_weeks_top5_friendly <- format(total_weeks_top5, big.mark = ",")
```

**Analysis and Growth Insights:**

Khakee: The Bihar Chapter, the top Hindi-language program, spent `r khakee_weeks_friendly` cumulative weeks in the Top 10 in India alone. To add to this notable statistic, the top 5 Hindi programs combined have garnered over `r total_weeks_top5_friendly` cumulative weeks in the Top 10. If one assumes that an average viewer sits through one show per week, this indicates an estimated Netflix audience of around `r estimated_customers_friendly` viewers (in India). It is worth mentioning, however, that several of these programs did not chart in the US. This demonstrates Netflix’s ability to produce region-specific hits. Growth trends in the long-term seem to suggest that India’s audience for Hindi-language content will continue to expand. This in turn will provide Netflix with great opportunities for subscriber retention, acquisition, and cultural impact.

**Conclusion:** 

It can be seen that the Hindi-language catalog on Netflix is proving to be a significant driver of engagement across India, positioning Netflix for continued growth and success in one of the world's fastest and largest-growing media markets.

## Press Release 3: Global TV Outperforms Films by Miles

**TV Shows Lead the Way: Netflix’s International Audience Demonstrates a Strong Preference for TV Shows**

As shown in the graph below, Netflix seems to dominate worldwide entertainment. Audiences are showing a preference towards serialized TV content. TV shows across multiple countries are consistently making the Top 10 weekly rankings, showcasing engagement trends globally as well as opportunities for content investment.

```{r, echo=TRUE}
#| code-fold: true
library(dplyr)
library(ggplot2)
library(stringr)
library(scales)

# Step 1: Prepare combined global data
global_content <- COUNTRY_TOP_10_CLEAN %>%
  filter(!is.na(Week)) %>%  # remove missing weeks
  left_join(
    GLOBAL_TOP_10 %>% select(show_title, category) %>% distinct(),
    by = c("Show_Title" = "show_title"),
    relationship = "many-to-many"
  )

# Step 2: Compute cumulative weeks in Top 10 by broad genre
genre_trends <- global_content %>%
  group_by(category) %>%
  summarise(
    total_weeks = sum(Cumulative_weeks_In_Top_10, na.rm = TRUE),
    n_countries = n_distinct(Country),
    .groups = "drop"
  ) %>%
  arrange(desc(total_weeks))

# Step 3: Highlight top genre
genre_trends$highlight <- ifelse(genre_trends$total_weeks == max(genre_trends$total_weeks),
                                 "Top Genre", "Other")
genre_trends$highlight <- factor(genre_trends$highlight, levels = c("Top Genre", "Other"))

# Step 4: Plot top genres globally
ggplot(genre_trends, aes(x = reorder(category, total_weeks),
                         y = total_weeks,
                         fill = highlight)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Top Genre" = "red", "Other" = "grey70")) +
  scale_y_continuous(labels = comma) +  # human-readable y-axis
  labs(
    title = "Top Content Types Globally on Netflix",
    subtitle = "Cumulative Weeks in Top 10 by Category",
    x = "Category",
    y = "Cumulative Weeks in Top 10",
    fill = "Highlight"
  ) +
  theme_minimal(base_size = 12)

# Step 5: Compute friendly numbers for markdown
top_category <- genre_trends$category[genre_trends$highlight == "Top Genre"]
total_weeks_top <- genre_trends$total_weeks[genre_trends$highlight == "Top Genre"]
countries_top <- genre_trends$n_countries[genre_trends$highlight == "Top Genre"]

total_weeks_top_friendly <- format(total_weeks_top, big.mark = ",")
```
## Analysis and Growth Insights

Netflix’s global audience shows a strong preference for the top content type:

**`r top_category`** spent **`r total_weeks_top_friendly`** cumulative weeks in the Top 10 across **`r countries_top`** countries.

One can see that Netflix has an ability to put out appealing content to international viewers. Although both English and Non-English Films perform well in regional markets, Non-English TV remains on top in terms of global engagement. This could indicate to investors and producers alike that Non-English serialized TV continues to be a key driver of viewership worldwide.

To add to this, growth trends in the longer-term suggest that focusing on serialized TV appealing to international markets can help in maximizing Netflix's audience engagement and retention, further exploiting expansion opportunities across diverse markets.
