---
title: Mini-Project 03 - Visualizing and Maintaining the Green Canopy of NYC
format: html
---

**Step 1:** Data Acquisition:

For the data relating to district boundary, I downloaded the official City Counci shapefile manually from the NYC Department of Planning website and saved it as a ZIP file into my data/mp03/ directory. Then, on R, I checked for the existing ZIP file for unzipping if needed. I then read the shapefile using sf::st_read() command, which converts the geometry of the file to a format called "WGS84" in order to ensure alignment with the tree data. This helps to provide a clean and reproducible workflow for loading district boundaries without the need for duplicated files.

```{r, echo=TRUE}

#| code-fold: true

get_nyc_council_districts <- function(
  dest_dir = "data/mp03",
  zip_name = "council_districts.zip"
) {
  # Path to zip file
  zip_path <- file.path(dest_dir, zip_name)

  if (!file.exists(zip_path)) {
    stop("ZIP file not found. Place it in data/mp03/ as council_districts.zip")
  }

  # Unzip only if needed
  shp_files <- list.files(dest_dir, pattern = "\\.shp$", full.names = TRUE)
  if (length(shp_files) == 0) {
    unzip(zip_path, exdir = dest_dir)
    shp_files <- list.files(dest_dir, pattern = "\\.shp$", full.names = TRUE)
  }

  if (length(shp_files) == 0) {
    stop("No .shp file found after unzipping. Check ZIP contents.")
  }

  # Read shapefile
  dist_sf <- sf::st_read(shp_files[1], quiet = TRUE)

  # Transform to WGS84 for mapping
  sf::st_transform(dist_sf, crs = "WGS84")
}

# Usage
nyc_council_dists <- get_nyc_council_districts()
```

**NYC Tree Points**

In order to obtain the entire NYC Street Tree dataset, I used the NYC OpenData SODA2 API in order to return data in paged batches. Since a single request only helps to retrieve the first 1,000 rows, my code adjust for the $limit and $offset parameters to 'page through' the entire dataset in order to save each chunk locally in my root data folder. Responsible API usage is ensured in this process and also allows me to obtain the entire set of tree points for spatial analysis in the subsequent sections.

```{r, echo=TRUE}

#| code-fold: true

library(httr2)
library(jsonlite)
library(dplyr)
library(purrr)
library(sf)

get_nyc_tree_points <- function(
  base_url = "https://data.cityofnewyork.us/resource/hn5i-inap.geojson",
  dest_dir = "data/mp03/trees",
  limit = 5000   
) {
  # Creates directory if needed
  if (!dir.exists(dest_dir)) dir.create(dest_dir, recursive = TRUE)

  offset <- 0
  all_files <- c()
  batch_num <- 1

  repeat {
    message("Downloading batch ", batch_num, " with offset = ", offset)

    # Building paged API request
    req <- request(base_url) |>
      req_url_query(`$limit` = limit, `$offset` = offset)

    # Performing request
    resp <- req_perform(req)

    # Parsing JSON as text then convert to R list
    content <- resp_body_string(resp)
    json_data <- fromJSON(content)

    # Saving raw JSON to file
    file_path <- file.path(dest_dir, paste0("trees_batch_", batch_num, ".geojson"))
    writeLines(content, file_path)
    all_files <- c(all_files, file_path)

    # Number of rows returned
    n_returned <- length(json_data$tree_id)

    # Stop if last batch
    if (n_returned < limit) break

    # Otherwise continue to next batch
    offset <- offset + limit
    batch_num <- batch_num + 1
  }

  message("Download complete. Reading and combining all batches...")

  # Read all GeoJSON batches into sf objects and combine
  tree_sf <- map_df(all_files, ~ st_read(.x, quiet = TRUE))

  # Return final sf object
  tree_sf
}

nyc_trees <- get_nyc_tree_points()
nyc_trees
```

**Step 2:** R Code for Task 2 (uses httr2, paging, caching, st_read, bind_rows)

```{r, echo=TRUE}

#| code-fold: true
#| label: setup-libs
#| message: false
#| warning: false
library(httr2)
library(sf)
library(dplyr)

get_nyc_tree_points <- function(
  base_url   = "https://data.cityofnewyork.us/resource/hn5i-inap.geojson",
  dest_dir   = "data/mp03",
  limit      = 2000,
  max_batches = Inf
) {
  if (!dir.exists(dest_dir)) {
    dir.create(dest_dir, recursive = TRUE)
  }

  offset     <- 0
  batch_idx  <- 1
  trees_list <- list()

  repeat {
    if (batch_idx > max_batches) {
      message("Reached max_batches = ", max_batches, ". Stopping early.")
      break
    }

    file_path <- file.path(
      dest_dir,
      sprintf("tree_points_%03d.geojson", batch_idx)
    )

    # Download only if this batch file doesn't already exist
    if (!file.exists(file_path)) {
      message("Downloading batch ", batch_idx,
              " (offset = ", offset, ", limit = ", limit, ")")

      req <- request(base_url) |>
        req_url_query(`$limit` = limit, `$offset` = offset)

      resp <- tryCatch(
        req_perform(req),
        error = function(e) {
          warning("Request failed for batch ", batch_idx, ": ",
                  conditionMessage(e))
          return(NULL)
        }
      )

      if (is.null(resp)) {
        message("Stopping because of repeated request errors.")
        break
      }

      writeBin(resp_body_raw(resp), file_path)
    } else {
      message("Using cached file: ", basename(file_path))
    }

    # Read this batch as sf from GeoJSON
    trees_batch <- sf::st_read(file_path, quiet = TRUE)

    # 🔧 Normalize ALL datetime columns to character so bind_rows won't choke
    dt_cols <- vapply(trees_batch, inherits, logical(1), what = "POSIXt")
    if (any(dt_cols)) {
      trees_batch[, dt_cols] <- lapply(trees_batch[, dt_cols, drop = FALSE], as.character)
    }

    trees_list[[batch_idx]] <- trees_batch

    n_returned <- nrow(trees_batch)

    if (n_returned == 0 || n_returned < limit) {
      message("Final batch size = ", n_returned, ". Finished paging.")
      break
    }

    offset    <- offset + limit
    batch_idx <- batch_idx + 1
  }

  # Final safety pass: making sure any leftover POSIX columns are characters
  trees_list <- lapply(trees_list, function(x) {
    dt_cols <- vapply(x, inherits, logical(1), what = "POSIXt")
    if (any(dt_cols)) {
      x[, dt_cols] <- lapply(x[, dt_cols, drop = FALSE], as.character)
    }
    x
  })

  dplyr::bind_rows(trees_list)
}

nyc_trees <- get_nyc_tree_points()
nyc_trees

saveRDS(nyc_trees, "data/mp03/nyc_trees.rds")
```

**Date Integration & Initial Exploration**

I made use of the NYC OpenData SODA2 API in order to download the full Street Tree dataset in paged GeoJSON batches. I saved each query result to the folder titled "data/mp03/ ", keeping consistency in the filename pattern. The function I created makes sure to only download a batch if the file is not already present in the data folder. It then reads all the batches with sf::st_read(), combining them into a single sf object which contains about 1.7 million tree points.

**Task 3**: Plot all Points

```{r, echo=TRUE}

#| code-fold: true

library(ggplot2)

ggplot() +
  geom_sf(data = nyc_council_dists, fill = "white", color = "black", size = 0.2) +
  geom_sf(data = nyc_trees, color = "darkgreen", alpha = 0.3, size = 0.1) +
  labs(
    title = "NYC Street Trees",
    subtitle = "All tree points plotted over City Council Districts",
    caption = "Source: NYC OpenData"
  ) +
  theme_minimal()
```
Spatially joining the various tree points to polygons in order for each tree to get a district ID.

```{r, echo=TRUE}

#| code-fold: true

nyc_trees_joined <- st_join(
  nyc_trees,
  nyc_council_dists |> dplyr::select(CounDist),
  join = st_within
)
```

**Task 4**: District Level Analysis of Tree Coverage

1) District with most trees

```{r, echo=TRUE}

#| code-fold: true
#| 
trees_by_count <- nyc_trees_joined |>
  count(CounDist) |>
  arrange(desc(n))

trees_by_count |>
  slice_max(n, n = 1)
```
The district with the most trees is District 51 (Staten Island) with 99,335 trees.

2) Council with highest density of trees

```{r, echo=TRUE}

#| code-fold: true

# add area in square km to district sf
nyc_council_dists <- nyc_council_dists |>
  mutate(area_sqkm = Shape_Area / 1e6)

density_by_district <- nyc_trees_joined |>
  st_drop_geometry() |>
  filter(!is.na(CounDist)) |>
  count(CounDist, name = "n_trees") |>
  left_join(
    nyc_council_dists |>
      st_drop_geometry() |>
      select(CounDist, area_sqkm),
    by = "CounDist"
  ) |>
  mutate(tree_density = n_trees / area_sqkm) |>
  arrange(desc(tree_density))

# top district by density
density_by_district |> slice_max(tree_density, n = 1)
```

The District with the highest density of trees is District 39 (Brooklyn) with approximately 496 trees per square kilometer of district area.

3) Which district has highest fraction of dead trees out of all trees?

```{r, echo=TRUE}

#| code-fold: true

dead_fraction <- nyc_trees_joined |>
  st_drop_geometry() |>
  filter(!is.na(CounDist)) |>
  mutate(is_dead = tpcondition == "Dead") |>
  group_by(CounDist) |>
  summarise(
    total_trees = n(),
    dead_trees  = sum(is_dead, na.rm = TRUE),
    frac_dead   = dead_trees / total_trees,
    .groups = "drop"
  ) |>
  arrange(desc(frac_dead))

# top district by fraction dead
dead_fraction |> slice_max(frac_dead, n = 1)
```
The district with the highest fraction of dead trees is District 32 (South Queens / Rockaway) with approximately 13.49% dead trees.

4) What is the most common tree species in Manhattan?

```{r, echo=TRUE}

#| code-fold: true
manhattan_species <- nyc_trees_joined |>
  st_drop_geometry() |>
  filter(CounDist %in% 1:10) |>
  count(genusspecies, sort = TRUE)

# top species
manhattan_species |> slice_max(n, n = 1)
```
The most common tree species in Manhattan is the "Thornless honeylocust". There are approximately 20,494 of them.

5) What is the species of the tree closest to Baruch's campus?

```{r, echo=TRUE}

#| code-fold: true

# Baruch College approx coordinates (lon, lat)
baruch <- st_sfc(
  st_point(c(-73.9830, 40.7403)),
  crs = 4326
)

# distance from each tree to Baruch (in meters)
nyc_trees_joined$dist_to_baruch <- as.numeric(st_distance(nyc_trees_joined, baruch))

nearest_tree <- nyc_trees_joined |>
  slice_min(dist_to_baruch, n = 1)

# look at key info for the nearest tree
nearest_tree |>
  st_drop_geometry() |>
  select(CounDist, genusspecies, tpcondition, dbh, dist_to_baruch)
```
The name of the species of trees closest to Baruch's campus is the "Callery Pear".

**Government Project Design**

Creating district specific-dataset

```{r, echo=TRUE}

#| code-fold: true

my_district <- 51

district_trees <- nyc_trees_joined |>
  filter(CounDist == my_district)

district_shape <- nyc_council_dists |>
  filter(CounDist == my_district)
```

Zoomed-In Map of Trees in District 51

```{r, echo=TRUE}

#| code-fold: true

ggplot() +
  geom_sf(data = district_shape, fill = "gray95", color = "black") +
  geom_sf(data = district_trees, aes(color = tpcondition), size = 0.5, alpha = 0.5) +
  scale_color_manual(values = c(
    "Excellent" = "darkgreen",
    "Good"      = "forestgreen",
    "Fair"      = "goldenrod",
    "Poor"      = "orange",
    "Dead"      = "red",
    "Critical"  = "purple",
    "Unknown"   = "gray"
  )) +
  labs(
    title = "Street Trees in NYC Council District 51",
    subtitle = "Colored by Condition (tpcondition)",
    color = "Tree Condition"
  ) +
  theme_minimal()
```
Quantitative Comparisons

```{r, echo=TRUE}

#| code-fold: true

comparison_table <- nyc_trees_joined |>
  st_drop_geometry() |>
  mutate(is_dead = tpcondition == "Dead") |>
  group_by(CounDist) |>
  summarise(
    total_trees = n(),
    dead_trees  = sum(is_dead, na.rm = TRUE)
  ) |>
  left_join(
    nyc_council_dists |> 
      mutate(area_sqkm = Shape_Area / 1e6) |> 
      st_drop_geometry(),
    by = "CounDist"
  ) |>
  mutate(tree_density = total_trees / area_sqkm)

compare_dists <- c(39, 32, 19, 51)  # example set

#Comparing three other districts
comparison_table |> 
  filter(CounDist %in% compare_dists)
```
Bar Chart

Dead Tree Comparison
```{r, echo=TRUE}

#| code-fold: true

comparison_table |>
  filter(CounDist %in% compare_dists) |>
  ggplot(aes(x = factor(CounDist), y = dead_trees, fill = factor(CounDist))) +
  geom_col() +
  labs(
    title = "Dead Trees by District",
    x = "Council District",
    y = "Number of Dead Trees",
    fill = "District"
  ) +
  theme_minimal()
```

Map Based Comparison

```{r, echo=TRUE}

#| code-fold: true

district_compare <- nyc_council_dists |> 
  filter(CounDist %in% c(51, 39))

trees_compare <- nyc_trees_joined |> 
  filter(CounDist %in% c(51, 39) & tpcondition == "Dead")

ggplot() +
  geom_sf(data = district_compare, fill = "gray92") +
  geom_sf(data = trees_compare, color = "red", size = 0.5, alpha = 0.6) +
  facet_wrap(~CounDist) +
  labs(
    title = "Dead Trees Comparison: D51 vs D39",
    subtitle = "Red points: trees listed as 'Dead'",
    x = "", y = ""
  ) +
  theme_minimal()
```

## NYC Parks Proposal: District 51 Program for Tree Revitalization

Being a staffer for the NYC Council Member representing **District 51** (Staten Island), I would like to propose a  **street-tree revitalization program** in order to strengthen and restore the urban canopy located in Staten Island (South Shore). District 51 possesses the **largest total number of street trees in NYC with approximately 91,335 trees**, however, large portions of trees are in **poor**, **critical**, or **dead** condition. The conditions of these less healthy trees is particularly visible along key residential areas such as **Hylan Boulevard** and **Amboy Road**. The goal of this initiative is to **replant lost canopy**, **improve overall environmental resilience** and to **replace hazardous trees**.

### Proposed Scope of Work

- **Replace 2,000** critically rated or dead trees  
- **Plant 1,500** brand new trees in areas where there is sparse canopy  
- **Remove 1,000** existing tree stumps in order to make more space for plantings in the future  
- **Perform risk-mitigation work**, such as treatment and/or pruning, on **500** high-risk trees which was identified in the dataset for NYC Parks.  

A spatial analysis of the NYC OpenData reveals that **District 51 has the highest tree count in the city**, however, it also ranks high amongst districts with **dead-tree burden**. When one compares this with **District 39 in Brooklyn** and both **Districts 19 and 32 in Queens**, District 51 shows both a **greater number of failing trees** as well as **neighborhoods where canopy loss is accelerating**. This is particularly true in coastal zones that are vulnerable to salt exposure, severe weather events as well as storm surges. These quantitative differences show that any additional resources from NYC Parks that are allocated to District 51 should have a **considerable impact on public safety as well as long-term canopy health**.

Furthermore, a more zoomed-in map of District 51 brings to light the distribution of tree conditions, with various groupings of **critical** and **dead** trees that are visible around **Tottenville**, **Great Kills** and **Annadale**. A bar chart comparing dead-tree counts across various districts further showcases the **need for targeted intervention** in District 51.

Through the process of directing a focused revitalization program in District 51, NYC Parks will be able achieve notable improvements in **pedestrian comfort**, **neighborhood aesthetics** and **environmental resilience**. This program illustrates a clear and **data driven rationale** in order to prioritize the South Shore during the next phase of discretionary funding for urban forestry.
