---
title: "Mini-Project 04: Just the Fact(-Check)s, Ma'am!"
author: "Tanmay Sharma"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    theme: cosmo
    df-print: paged
    number-sections: true
    smooth-scroll: true
    code-fold: true
    code-overflow: wrap
    fig-cap-location: bottom
    tbl-cap-location: top
---

## Introduction

In this mini-project, I construct a reproducible pipeline to obtain and analyze
the U.S. Bureau of Labor Statistics (BLS) **Current Employment Statistics (CES)**
data for total nonfarm payrolls and their subsequent revisions (January 1979–
June 2025). Using web scraping with `httr2` and `rvest`, I download:

- The final seasonally-adjusted CES employment levels, and  
- The month-to-month revisions from the CES revisions page.

I then join these datasets into a single table for statistical exploration,
formal hypothesis testing, and fact-checking recent political claims about
BLS revisions and the firing of Commissioner Erika McEntarfer.

---

## Executive Summary

- **Data:** Monthly CES total nonfarm payroll levels and corresponding revisions, January 1979 – June 2025.  
- **Big picture:** Revisions are **tiny relative to total employment**, centered near zero, and do **not** show a clear partisan or post-2020 “rigging” pattern.  
- **Key tests:**  
  - One-sample t-tests and bootstrap CIs show average revisions are **not meaningfully different from zero**.  
  - Proportion tests and permutation tests show **only weak differences** in positive/negative revision rates over time.  
  - Revisions are **larger when the labor market is more volatile**, not when the political party changes.
- **Bottom line:** Claims that CES revisions were “rigged” to hurt Trump are **inconsistent** with the long-run data; revisions behave like **normal measurement noise** in a very large labor market.

---

## Data Acquisition

We obtain monthly **Total Nonfarm Payroll Employment** (seasonally adjusted) from the U.S. Bureau of Labor Statistics (BLS) using their PDQ interface, covering **January 1979 – June 2025**.

### Load Required Libraries

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

needed_pkgs <- c(
  "tidyverse",
  "httr2",
  "rvest",
  "lubridate",
  "infer",
  "janitor"
)

to_install <- setdiff(needed_pkgs, rownames(installed.packages()))
if (length(to_install) > 0) {
  install.packages(to_install)
}

invisible(lapply(needed_pkgs, library, character.only = TRUE))
```

## Task 1: Downloading CES Total Nonfarm Payroll**

We begin by loading the R packages required for downloading, cleaning, and analyzing the Current Employment Statistics (CES) dataset.  
If any packages are missing, they are installed automatically to ensure full reproducibility.

```{r task1_ces_download, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

library(tidyverse)
library(httr2)
library(rvest)
library(lubridate)
library(janitor)

# --- 1. Building the HTTP request to the PDQ page ---

ces_base <- "https://data.bls.gov/pdq/SurveyOutputServlet"

ces_req <- request(ces_base) |>
  req_method("POST") |>
  req_body_form(
    survey    = "ce",
    from_year = "1979",
    to_year   = "2025",
    # This is the CES series ID for All Employees, Total Nonfarm, SA
    series_id = "CES0000000001"
  )

# Performing request and getting HTML
ces_html <- ces_req |>
  req_perform() |>
  resp_body_html()

# --- 2. Grabbing the main data table from the HTML ---

ces_raw_tbl <- ces_html |>
  html_element("table") |>
  html_table(fill = TRUE) |>
  clean_names()

# Peeking at the raw table structure
head(ces_raw_tbl)

all_tables <- ces_html |> html_elements("table")
length(all_tables)

all_tables[[2]] |> html_table(fill = TRUE) |> head()

# Getting ALL tables from the HTML
all_tables <- ces_html |> html_elements("table")

# Extracting the SECOND table, (Year–Month table)
ces_data_raw <- all_tables[[2]] |>
  html_table(fill = TRUE) |>
  clean_names()

# Checking it
head(ces_data_raw)
```

### Clean and Reshape CES Data

Once the raw table is extracted, we convert it into a tidy monthly time-series structure.  
This includes standardizing column names, reshaping year–month columns into a single date variable, converting values to numeric format, and filtering to match the project time window (Jan 1979 – Jun 2025).

```{r task1_ces_clean, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

ces_levels <- ces_data_raw |>
  clean_names() |>
  pivot_longer(
    cols = -year,
    names_to  = "month",
    values_to = "level"
  ) |>
  mutate(
    month = str_to_title(month),
    date_str = paste(year, month),
    date = lubridate::ym(date_str),
    level = as.numeric(level)
  ) |>
  drop_na(date, level) |>
  filter(date <= as.Date("2025-06-01")) |>
  arrange(date) |>
  select(date, level)

head(ces_levels)
```

## Task 2 — Download CES Revision Tables

In addition to monthly payroll employment estimates, the BLS provides revision tables comparing the **initial** estimate with the **final** benchmarked value.  
We download these tables directly from the BLS website for the full period (1979–2025).

### Retrieve Revisions Page HTML

```{r task2_fetch_html, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

url_rev <- "https://www.bls.gov/web/empsit/cesnaicsrev.htm"

# Simpler: let rvest handle the download directly
rev_html <- rvest::read_html(url_rev)

rev_html
```

### Inspect Revision Table Structure

The revision page contains multiple separate HTML tables — one per year.  
Here, I verify the number of tables and inspect their ID patterns to confirm  
that each corresponds to a specific year of CES revisions.

```{r task2_inspect_tables, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

library(rvest)
library(tidyverse)
library(lubridate)

# All tables on the revisions page
rev_tables <- rev_html |> html_elements("table")

length(rev_tables)

# Looking at the first few table IDs to see the pattern
head(html_attr(rev_tables, "id"), 20)
```

### Extract Revisions for a Single Year

BLS publishes a separate revision table for each year.  
This helper function automatically locates the correct table by its HTML `id`,  
cleans the rows, and returns a tidy month-level dataset of original vs. final values.

```{r task2_extract_year_function_fixed, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

extract_ces_revisions_year <- function(year) {
  
  table_id <- as.character(year)
  
  tbl <- rev_html |>
    html_element(paste0("table#", table_id)) |>
    html_table(header = FALSE, fill = TRUE) |>
    janitor::clean_names()
  
  # Dropping the first 3 header rows, then keeping the 12 months
  tbl <- tbl |>
    slice(-(1:3)) |>
    slice(1:12) |>
    select(
      month    = 1,  # month name (Jan, Feb, ...)
      original = 3,  # first estimate
      final    = 4   # final (third) estimate
    ) |>
    mutate(
      month    = stringr::str_to_title(month),
      date_str = paste(year, month),
      date     = lubridate::ym(date_str),
      original = as.numeric(original),
      final    = as.numeric(final),
      revision = final - original
    ) |>
    select(date, original, final, revision) |>
    arrange(date)
  
  tbl
}

# Testing again on 2024
extract_ces_revisions_year(2024)
```

### Build Full CES Revisions Dataset

Using the helper function above, I loop through all available years (1979–2025)  
to compile a single tidy dataset of monthly revision values, filtered through June 2025.

```{r task2_build_full_revisions, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

library(purrr)
library(dplyr)

# All years with published revisions
years <- 1979:2025

ces_revisions <- map_dfr(years, extract_ces_revisions_year) |>
  # Only keeping months through June 2025 as per instructions
  filter(date <= as.Date("2025-06-01")) |>
  arrange(date)

# Quick check
ces_revisions |> dplyr::slice_head(n = 12)
ces_revisions |> dplyr::slice_tail(n = 12)
```

### Join Employment Levels with Revision Data

Here, I merge the employment levels (Task 1) with revision data (Task 2)  
into a unified monthly dataset and create additional features for analysis.

```{r task3_join_levels_revisions, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

library(dplyr)
library(lubridate)

ces_full <- ces_levels |>          # from Task 1
  left_join(ces_revisions, by = "date") |>
  mutate(
    year       = year(date),
    month_num  = month(date),
    month_name = month(date, label = TRUE, abbr = TRUE),
    abs_revision     = abs(revision),
    rel_revision     = revision / level,
    abs_rel_revision = abs(revision) / level
  )

ces_full |> slice_head(n = 12)
```

## Task 3 — Data Exploration & Visualization

To understand the overall behavior of CES employment and revisions over the long run,  
I compute six key descriptive statistics across the full sample (1979–2025).

```{r task3_stats, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

library(dplyr)

ces_summary <- ces_full |>
  summarise(
    n_months                 = n(),
    avg_level                = mean(level, na.rm = TRUE),
    sd_level                 = sd(level, na.rm = TRUE),
    avg_revision             = mean(revision, na.rm = TRUE),
    sd_revision              = sd(revision, na.rm = TRUE),
    avg_abs_revision         = mean(abs_revision, na.rm = TRUE),
    avg_rel_revision_pct     = mean(rel_revision, na.rm = TRUE) * 100,
    pct_negative_revisions   = mean(revision < 0, na.rm = TRUE) * 100
  )

knitr::kable(
  ces_summary,
  caption = "Summary statistics for CES levels and revisions, 1979–2025.",
  digits = 2
)
```

### Visualization 1 — CES Revisions Over Time

This plot tracks monthly revisions (final − original estimate) from 1979–2025,  
showing how updates fluctuate above and below zero across economic cycles.

```{r task3_plot1_revisions_ts, echo=TRUE, message=FALSE, warning=FALSE}
#| fig-width: 7
#| fig-height: 4
#| code-fold: true

library(ggplot2)

ggplot(ces_full, aes(date, revision)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "CES Revisions Over Time (1979–2025)",
    subtitle = "Final minus original sector-level employment change",
    x = "Date",
    y = "Revision (thousands of jobs)"
  )
```

### Visualization 2 — Absolute Revisions Over Time

This view focuses on the *magnitude* of revisions — how large the corrections are,  
regardless of whether they increased or decreased employment levels.

```{r task3_plot2_absrev_ts, echo=TRUE, message=FALSE, warning=FALSE}
#| fig-width: 7
#| fig-height: 4
#| code-fold: true

ggplot(ces_full, aes(date, abs_revision)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Absolute CES Revisions Over Time",
    x = "Date",
    y = "Absolute Revision (|final - original|)"
  )
```

### Visualization 3 — Revisions as % of Employment Level

This visualization scales revisions relative to total employment,  
showing how **significant** changes are in context — even large job changes  
often represent **very small percentages** of the labor market.


```{r task3_plot3_relrev_ts, echo=TRUE, message=FALSE, warning=FALSE}
#| fig-width: 7
#| fig-height: 4
#| code-fold: true

ggplot(ces_full, aes(date, rel_revision * 100)) +
  geom_line(color = "firebrick") +
  labs(
    title = "Revisions Relative to Employment Level (%)",
    x = "Date",
    y = "Revision as % of Level"
  )
```

### Visualization 4 — Distribution of CES Revisions

This histogram displays how revisions are distributed over the entire sample period.  
Most revisions cluster close to **zero**, with a relatively small number of  
**large upward or downward adjustments**, consistent with typical survey variability.

```{r task3_plot4_revision_hist, message=FALSE, echo=TRUE, warning=FALSE}
#| fig-width: 7
#| fig-height: 4
#| code-fold: true

ggplot(ces_full, aes(revision)) +
  geom_histogram(bins = 40, color = "white", fill = "darkblue") +
  labs(
    title = "Distribution of CES Revisions (1979–2025)",
    x = "Revision (thousands of jobs)",
    y = "Count"
  )
```

## Task 4 — Statistical Inference Setup

Before running formal hypothesis tests, I create several helper variables on the  
joined CES dataset. These include indicators for **negative revisions**,  
**post-2000** and **post-2020** periods, the **absolute** and **relative** size  
of revisions, and the **monthly change in employment levels**. These derived  
variables are used throughout the tests in Task 4.

```{r task4_setup, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

library(dplyr)
library(infer)
library(lubridate)

# Adding all needed variables to ces_full
ces_full <- ces_full |>
  arrange(date) |>
  mutate(
    year              = year(date),
    negative_revision = revision < 0,
    post_2000         = year >= 2000,
    abs_revision      = abs(revision),
    rel_revision      = revision / level,
    abs_rel_revision  = abs(revision) / level,
    big_1pct          = abs_rel_revision > 0.01,      # >1% of level
    post_2020         = year >= 2020,
    change            = level - lag(level),
    abs_change        = abs(change)
  )
```

### 4.1 — Has the fraction of negative revisions increased post-2000?

To evaluate whether CES revisions have become **more negative** in the modern era,  
I compare the proportion of downward revisions **before vs. after 2000**.  
A **two-sample proportion test** determines whether these differences are  
statistically significant.

```{r task4_q1_prop_negative_post2000, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

# Counting negative vs total before/after 2000
neg_tab <- ces_full |>
  filter(!is.na(negative_revision), !is.na(post_2000)) |>
  group_by(post_2000) |>
  summarise(
    neg = sum(negative_revision),
    n   = n(),
    .groups = "drop"
  )

neg_tab

# Order: FALSE = pre-2000, TRUE = post-2000
x_neg <- neg_tab$neg[order(neg_tab$post_2000)]
n_neg <- neg_tab$n[order(neg_tab$post_2000)]

# Base R two-sample test for equality of proportions
prop_neg_post2000 <- prop.test(
  x = x_neg,
  n = n_neg,
  alternative = "two.sided",
  correct = FALSE
)

prop_neg_post2000
```

#### 4.2 — Has the fraction of large (>1%) revisions increased post-2020?

To check whether revisions have become *unusually large* in recent years,  
I flag months where the revision is more than 1% of employment:

- `big = 1` if `|revision / level| > 0.01`
- `big = 0` otherwise

In my CES sample (Jan 1979–Jun 2025), **no months** exceed this 1% threshold in either
the pre-2020 or post-2020 period. As a result, the two-sample proportion test has
`prop 1 = 0`, `prop 2 = 0` and returns `NaN` for the test statistic and p-value.

Substantively, this means that revisions larger than 1% of total employment are
**essentially non-existent** in the data, even during the volatile post-2020 period.
So while revisions can occasionally look large in absolute job counts,
they almost never reach the “>1% of employment” threshold.

```{r task4_q2_prop_big1pct_post2020, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
big_tab <- ces_full |>
  filter(!is.na(big_1pct), !is.na(post_2020)) |>
  group_by(post_2020) |>
  summarise(
    big = sum(big_1pct),
    n   = n(),
    .groups = "drop"
  )

big_tab

# Order: FALSE = pre-2020, TRUE = post-2020
x_big <- big_tab$big[order(big_tab$post_2020)]
n_big <- big_tab$n[order(big_tab$post_2020)]

prop_big_post2020 <- prop.test(x = x_big, n = n_big)
prop_big_post2020
```

### 4.3 — Is the average revision significantly different from zero?

Because revisions represent corrections to initially reported employment figures,  
we expect them to **average out near zero** over time if the BLS estimates are  
unbiased.  
A **one-sample t-test** formally evaluates the null hypothesis:  

> **H₀:** Mean revision = 0  
> (no systematic upward or downward bias)

```{r task4_q3_mean_revision_zero, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

# One-sample t-test using infer::t_test: H0: mean revision = 0

ttest_mean_rev <- ces_full |>
  drop_na(revision) |>
  t_test(revision ~ NULL, mu = 0)

ttest_mean_rev
```

### 4.4 — Has the average revision increased post-2020?

The pandemic era brought heightened labor market volatility and lower survey response rates.  
To assess whether this led to **systematically larger revisions**, we compare the mean revision:

- **Before 2020**
- **After 2020**

A **two-sample t-test** evaluates whether the post-2020 average revision is significantly higher:

> **H₀:** Mean revision (post-2020) = Mean revision (pre-2020)  
> **H₁:** Mean revision (post-2020) > Mean revision (pre-2020)

```{r task4_q4_mean_revision_post2020, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
ttest_mean_rev_post2020 <- ces_full |>
  drop_na(revision, post_2020) |>
  t_test(
    revision ~ post_2020,
    order = c("FALSE", "TRUE")   # estimate = post-2020 − pre-2020
  )

ttest_mean_rev_post2020
```

### 4.5 — Are revisions larger when the underlying change is larger?

Revisions may grow when the labor market is more volatile — for example,  
during recessions or sudden employment shocks.  
To examine this relationship, we measure the **correlation** between:

- Absolute change in employment level (`|Δ level|`)
- Absolute revision (`|final − original|`)

A positive and significant correlation would suggest that  
**larger real movements** lead to **larger revisions**, indicating  
measurement noise rather than systemic manipulation.

```{r task4_q5_cor_abschange_absrevision, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
ces_cor <- ces_full |>
  drop_na(abs_change, abs_revision)

cor_abs <- cor.test(ces_cor$abs_change, ces_cor$abs_revision)

cor_abs
```

## Task 5 — Fact-Checking Public Claims About CES Revisions

This final section applies the results from Tasks 3 and 4 to evaluate **two real-world claims** made about employment revisions in the U.S. Current Employment Statistics (CES) program — particularly regarding the accuracy and transparency of the BLS during the period surrounding the firing of Commissioner Erika McEntarfer.

Each fact-check includes:

- Relevant **summary statistics** from the CES dataset  
- At least **two visualizations** from Task 3  
- Appropriate **hypothesis testing** (with parallels to Task 4)  
- A final judgment using a **Politifact-style Truth Scale**  
  (ranging from **True** to **Pants on Fire**)

The CES dataset used here spans **January 1979 – June 2025**, providing a long-run perspective on revision behavior across economic cycles and presidential administrations.

```{r task5_setup, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
library(dplyr)
library(lubridate)
```

### Basic Summary Statistics for CES Levels and Revisions

```{r task5_stats_basic, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
ces_summary_5 <- ces_full |>
summarise(
n_months = n(),
avg_level = mean(level, na.rm = TRUE),
sd_level = sd(level, na.rm = TRUE),
avg_revision = mean(revision, na.rm = TRUE),
sd_revision = sd(revision, na.rm = TRUE),
avg_abs_revision = mean(abs(revision), na.rm = TRUE),
avg_rel_revision_pct = mean(revision / level, na.rm = TRUE) * 100,
pct_negative_revisions = mean(revision < 0, na.rm = TRUE) * 100
)

knitr::kable(
  ces_summary_5,
  caption = "Summary statistics for CES levels and revisions used in fact-checking.",
  digits = 2
)
```

### Largest and Smallest Revisions in the Sample

```{r task5_stats_extremes, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
max_rev_5 <- ces_full |>
summarise(
max_revision = max(revision, na.rm = TRUE),
min_revision = min(revision, na.rm = TRUE),
max_abs_revision = max(abs(revision), na.rm = TRUE)
)

max_rev_5
```

### Pre-2000 vs Post-2000 Share of Negative Revisions (Proportion Test)

```{r task5_prop_neg_prepost2000, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
ces_neg_tmp <- ces_full |>
mutate(
year = year(date),
post_2000 = year >= 2000,
negative_revision = revision < 0
)

neg_tab_5 <- ces_neg_tmp |>
filter(!is.na(negative_revision), !is.na(post_2000)) |>
group_by(post_2000) |>
summarise(
neg = sum(negative_revision),
n = n(),
.groups = "drop"
)

neg_tab_5
knitr::kable(
  neg_tab_5,
  caption = "Share of negative revisions before and after 2000."
)

x_neg_5 <- neg_tab_5$neg[order(neg_tab_5$post_2000)]
n_neg_5 <- neg_tab_5$n[order(neg_tab_5$post_2000)]

prop_neg_test_5 <- prop.test(x = x_neg_5, n = n_neg_5)
prop_neg_test_5
```

### One-Sample t-Test: Is the Mean Revision Different from Zero?

```{r task5_ttest_mean_revision, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

rev_vec_5 <- ces_full$revision
rev_vec_5 <- rev_vec_5[!is.na(rev_vec_5)]

t_mean_5 <- t.test(rev_vec_5, mu = 0)
t_mean_5
```

### Big Revisions (>1% of Level) — Pre- vs Post-2020 (Proportion Test)

```{r task5_prop_big1pct_prepost2020, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
ces_big_tmp <- ces_full |>
mutate(
abs_rel_revision = abs(revision / level),
big1pct = abs_rel_revision > 0.01,
year = year(date),
post_2020 = year >= 2020
)

big_tab_5 <- ces_big_tmp |>
filter(!is.na(big1pct), !is.na(post_2020)) |>
group_by(post_2020) |>
summarise(
big = sum(big1pct),
n = n(),
.groups = "drop"
)

big_tab_5

x_big_5 <- big_tab_5$big[order(big_tab_5$post_2020)]
n_big_5 <- big_tab_5$n[order(big_tab_5$post_2020)]

big1pct_test_5 <- prop.test(x = x_big_5, n = n_big_5)
big1pct_test_5
```

### Relationship Between Absolute Level Change and Absolute Revision (Correlation Test)

```{r task5_abschange_absrev_cor, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
ces_cor_5 <- ces_full |>
arrange(date) |>
mutate(
change = level - lag(level),
abs_change = abs(change),
abs_revision = abs(revision)
) |>
filter(!is.na(abs_change), !is.na(abs_revision))

cor_test_result_5 <- cor.test(ces_cor_5$abs_change, ces_cor_5$abs_revision)
cor_test_result_5
```

### Do Large-Change Months Have Larger Revisions? (Two-Sample t-Test)

```{r task5_largechange_ttest, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
q75_change_5 <- quantile(ces_cor_5$abs_change, 0.75, na.rm = TRUE)

ces_cor2_5 <- ces_cor_5 |>
mutate(
large_change = abs_change >= q75_change_5
)

large_change_test_5 <- t.test(abs_revision ~ large_change, data = ces_cor2_5)
large_change_test_5
```

### Fact Check 1 – Trump: “The numbers were rigged” and revised down by ~900,000

In early August 2025, President Donald Trump publicly claimed that BLS jobs numbers were **“rigged”** and that they had been revised **“down by almost 900,000 jobs”** after the 2024 election, describing this as the “biggest revision in history” and a justification for firing BLS Commissioner Erika McEntarfer.

This claim has two parts:

1. That recent CES revisions are extraordinarily large and one-sided  
2. That these revisions are politically manipulated (“rigged”).

#### What do revisions look like in the CES data?

Across my full sample (January 1979–June 2025) there are 558 monthly observations.
The **average CES employment level** is about **124,707 thousand jobs**  
(roughly **125 million jobs**), with a standard deviation of about **19,994 thousand**.
This is a very large labor market.

For revisions (final – original):

- The **average revision** is `r round(ces_summary_5$avg_revision, 1)` thousand jobs, with standard deviation `r round(ces_summary_5$sd_revision, 1)` thousand.
- The **average absolute revision** is `r round(ces_summary_5$avg_abs_revision, 1)` thousand jobs.
- As a **percentage of employment**, the average revision is only about `r round(ces_summary_5$avg_rel_revision_pct, 3)`% of the level.
- About `r round(ces_summary_5$pct_negative_revisions, 1)`% of revisions are negative; downward revisions are common, but not overwhelmingly dominant.

Figure 1 from Task 3 (**“CES Revisions Over Time (1979–2025)”**) shows that revisions fluctuate both above and below zero across many administrations. Figure 4 (**“Distribution of CES Revisions (1979–2025)”**) shows a roughly symmetric distribution around zero, with most revisions fairly close to zero and a relatively small number of extreme outliers.

### Largest Monthly Revision in the Sample

```{r task5_print_maxrev, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
max_rev_5
```

This is large in absolute terms, but it happens in the context of a very large labor market and is not unique to one president or one year.

### Are Revisions Systematically More Negative Post-2000?

This provides context for political claims that BLS revisions intentionally push job numbers downward in recent years.  
We compare the **share of negative revisions** before vs after 2000.

```{r task5_neg_post2000_test, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
neg_tab_5
prop_neg_test_5
```

The table neg_tab_5 shows the fraction of months with negative revisions before and after 2000, and prop_neg_test_5 is a formal two-sample proportion test. While there may be some difference across periods, the results do not show a dramatic shift to predominantly negative revisions that would match the idea of “rigged” numbers under a particular administration.

### Is the Average Revision Far From Zero? (One-Sample t-Test)

This test checks whether revisions are systematically biased upward or downward across the full sample — a core part of the “rigged numbers” narrative.

```{r task5_mean_rev_zero_test, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
t_mean_5
```
This one-sample t-test shows that the average revision is small in magnitude relative to both its own standard deviation and the level of employment. The confidence interval for the mean revision is narrow relative to total employment and close to zero in percentage terms. This
is consistent with revisions being noise around an unbiased estimate, not clear evidence of systemic manipulation in one direction.

Rating: Trump’s “rigged and revised down by 900,000” claim

The “almost 900,000” figure refers to an annual benchmark revision spread over many months, publicly documented in advance and not hidden. My month-by-month CES dataset shows:

Revisions are small on average relative to the overall employment
level. They are not consistently negative after 2000. Formal tests do not support a story of a large, one-sided shift in revisions that coincides with Trump’s political narrative. Given the data and the statistics above, the idea that the BLS jobs numbers were “rigged” to hurt Trump is not supported by the CES revisions.

Politifact-style rating for Claim 1: Pants on Fire

The claim significantly misrepresents both the size and the pattern of
CES revisions.

### Fact Check 2 — Wilcox / PIIE: “Challenges? Yes. Rigged data? No.”

Economist David Wilcox at the Peterson Institute for International Economics reviewed CES revisions and concluded that while measurement challenges exist — especially declining survey response rates — **there is no evidence of manipulation or “rigged” data**.

We evaluate a simplified version of his claim:

> **Claim:** CES revisions are not systematically biased or unusually large.  
> They represent normal measurement noise in a large labor market.

---

### How Large Are Revisions Relative to Total Employment?

From the summary statistics above:

- **Mean absolute revision:**  
  `r round(ces_summary_5$avg_abs_revision, 1)` thousand jobs  
- **Average employment level:**  
  `r round(ces_summary_5$avg_level, 0)` thousand jobs  
- **Mean revision as share of employment:**  
  `r round(ces_summary_5$avg_rel_revision_pct, 3)`%

These figures are very small relative to the overall labor market, indicating that typical revisions reflect **modest measurement updates**, not suspicious adjustments.

#### Are Large Revisions More Common in Recent Years?

Figures 2 and 3 show that revisions can spike during major economic disruptions —
such as recessions or the COVID period — but they generally remain well under 1% of
total employment.

To formally identify “big” revisions, I flagged any month where

\[
\left|\frac{\text{revision}}{\text{level}}\right| > 0.01
\]

In practice, **no months** in my sample exceed this 1% threshold, either before or
after 2020. The table `big_tab_5` therefore has `big = 0` in both periods, and the
two-sample proportion test in `big1pct_test_5` returns `NaN` (there is no variation
to compare).

Interpretation: revisions that are larger than 1% of total employment are
**extremely rare to non-existent** in the CES data, even in the post-2020 era.
This is consistent with normal measurement updates in a huge labor market,
not with unusually manipulated data.

```{r task5_big_revision_freq, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
big_tab_5
big1pct_test_5
```
Even during recent volatility, **large revisions remain rare**, offering no statistical support for claims of unusually manipulated data in the current era.

The table `big_tab_5` shows the frequency of >1% revisions before and after 2020, and  
`big1pct_test_5` provides a two-sample proportion test.  
While slightly more common post-2020, **large revisions remain rare overall** — consistent with normal measurement variation, not manipulation.

---

### Do Larger Employment Swings Lead to Larger Revisions?

A natural question:  
Are large revisions occurring because the labor market itself is moving more dramatically?

To test this, I examined the relationship between:

- |Change in employment (level − lag(level))|  
- |Revision (final − original)|

```{r task5_cor_change_revision, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
cor_test_result_5
```

The statistically significant **positive correlation** confirms that CES revisions tend to be larger when the labor market itself is more volatile — such as during recessions or the COVID period. This behavior is consistent with **normal survey noise**, not political tampering.

---

### Do Large Employment Shifts Cause Larger Revisions?

To further check this idea, I separated months into:

- **Large-change months** → top 25% of absolute employment changes  
- **Typical months** → the remaining 75%

Then I tested whether large-change months have bigger revisions:

```{r task5_large_change_ttest, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
large_change_test_5
```

This two-sample t-test shows that **large-change months** have significantly larger revisions, reinforcing the interpretation that revisions scale with true labor-market volatility — not with politics.

---

### Rating: Wilcox’s “Challenges? Yes. Rigged Data? No.” Claim

Putting all statistical evidence together:

- Revisions are **small** compared to the overall employment level.
- They remain **balanced around zero**, not consistently negative.
- Larger revisions appear when the **labor market is more volatile**, not when politics change.
- Revisions larger than **1% of employment essentially never occur** in the sample.

These findings support Wilcox’s position: there are measurement challenges,
but no clear signs of political manipulation in the CES revisions.

**Politifact-style Rating:** 🟢 *Mostly True*

CES revisions behave like **normal measurement noise** in a large, dynamic labor market — not rigged data.

## Extra Credit — 1️⃣ Non-Technical Explanation of Computational Inference

Traditional inference uses mathematical formulas to decide whether a result is statistically significant — for example, whether the **average revision differs from zero**. These formulas rely on assumptions about the data (like normality).

**Computationally intensive inference** flips the process:

- Instead of relying on formulas…
- We let the **computer simulate** many alternative versions of the data.

How it works:

- Use the real CES data as a **template**
- Repeatedly **resample** from it (bootstrap) or **shuffle group labels** (permutation test)
- Recalculate the statistic of interest each time
- Compare our real result to this **simulated distribution**

Why this is useful:

- Works even when theory-based assumptions might be shaky
- Provides **visual and intuitive** evidence about what is “extreme”
- Reflects the real structure and quirks of the observed data

In the context of this mini-project, these methods help us check:

> Are the revision patterns we see **unusual enough** to suggest rigging —
> or are they simply **within the normal range of sampling variation**?

Computers let us evaluate that question directly — **without** relying exclusively on theoretical formulas.

---

### 🔁 Schematic Workflow for Computational Inference

```{r extra_flowchart, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

library(ggplot2)
library(dplyr)

flow_steps <- tibble::tibble(
  step = c(
    "1. Pose a question\n(e.g., Is mean revision = 0?)",
    "2. Specify a null model\n(no effect / no difference)",
    "3. Resample the data\n(bootstrap or permute labels)",
    "4. Compute the statistic\nfor each resample",
    "5. Build simulated distribution\n(null or bootstrap)",
    "6. Compare observed statistic\nwith simulated distribution"
  ),
  x = 1,
  y = seq(6, 1, by = -1)
)

ggplot(flow_steps, aes(x = x, y = y, label = step)) +
  # drawing arrows first (behind boxes)
  geom_segment(
    aes(x = 1, xend = 1, y = y - 0.4, yend = y - 0.9),
    data = flow_steps |> filter(y > min(y)),
    arrow = arrow(length = unit(0.12, "inches")),
    linewidth = 0.4
  ) +
  # then drawing the boxes on top
  geom_label(size = 4.2, label.size = 0.4, fill = "white") +
  scale_x_continuous(NULL, breaks = NULL, limits = c(0.8, 1.2)) +
  scale_y_continuous(NULL, breaks = NULL, limits = c(0.2, 6.8)) +
  theme_minimal(base_size = 13) +
  ggtitle("Schematic: Computationally Intensive Inference Workflow")
```

This diagram summarizes the main idea:

1. Start with a real question and a null hypothesis.
2. Use resampling (bootstrap) or permutation to generate many “what if” datasets.
3. Compute the statistic each time and then compare our real statistic to that reference distribution.

### 🔬 Computational Tests for CES Revisions

We apply three resampling-based tests (bootstrap & permutation) using **infer** to evaluate:  
1️⃣ Mean revision (bootstrap mean)  
2️⃣ Median revision (bootstrap median)  
3️⃣ Difference in probability of positive revisions pre- vs post-2000 (permutation test)

```{r extra_comp_tests_setup, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
library(infer)
library(dplyr)
library(lubridate)
```

### 🧹 Prepare Clean Revision Data

We subset and clean the revision data to ensure it is ready for resampling-based inference.

```{r extra_clean_rev, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true

ces_rev <- ces_full |>
select(date, level, revision) |>
filter(!is.na(revision), !is.na(level))
```

### 📊 3.1 Bootstrap CI for the Mean Revision (Mean Analogue of a t-Test)

We first compute the mean revision observed in the actual CES sample.

```{r extra_obs_mean_revision, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
obs_mean_rev <- ces_rev |>
summarise(mean_rev = mean(revision)) |>
pull(mean_rev)

obs_mean_rev
```

#### 🔁 Bootstrap Distribution of the Mean Revision

We use bootstrap resampling to generate a simulated distribution of mean revisions under the assumption that the observed data reflect normal sampling variation.

```{r extra_boot_mean_revision, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
boot_mean_rev <- ces_rev |>
specify(response = revision) |>
generate(reps = 2000, type = "bootstrap") |>
calculate(stat = "mean")
```

#### 📏 95% Percentile Bootstrap Confidence Interval — Mean Revision

A bootstrap CI provides a **distribution-free** estimate of uncertainty around the mean revision.

```{r extra_boot_ci_mean, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
boot_ci_mean <- boot_mean_rev |>
get_confidence_interval(level = 0.95, type = "percentile")

boot_ci_mean
```

#### 🔹 Observed Median Revision (Bootstrap Median CI Setup)

We compute the median revision as a more **robust** measure (less sensitive to outliers than the mean).

```{r extra_obs_median_rev, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
obs_median_rev <- ces_rev |>
summarise(median_rev = median(revision)) |>
pull(median_rev)

obs_median_rev
```

#### 🔹 Bootstrap Distribution of the Median Revision

We resample from the observed data **2,000 times** to approximate the typical range of median revisions.

```{r extra_boot_median_rev, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
boot_median_rev <- ces_rev |>
specify(response = revision) |>
generate(reps = 2000, type = "bootstrap") |>
calculate(stat = "median")
```

#### 🔹 95% Percentile Bootstrap CI for Median Revision

This interval provides a robust estimate of the “typical” revision size, less influenced by extreme values than the mean.

```{r extra_boot_ci_median, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
boot_ci_median <- boot_median_rev |>
get_confidence_interval(level = 0.95, type = "percentile")

boot_ci_median
```

#### 🔹 Prepare Data for Permutation Test — Positive Revisions (Pre vs Post-2000)

We classify each month as having a **positive revision or not**, and divide the sample into **pre-2000 vs post-2000** periods.

```{r extra_perm_data_setup, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
ces_pos <- ces_full |>
  mutate(
    year         = year(date),
    post_2000    = year >= 2000,
    positive_rev = revision > 0
  ) |>
  filter(!is.na(post_2000), !is.na(positive_rev)) |>
  mutate(
    post_2000    = factor(post_2000, levels = c(FALSE, TRUE)),
    positive_rev = factor(positive_rev, levels = c(FALSE, TRUE))
  )
```

#### 🔹 Observed Difference in Positive Revision Rates  
*(Post-2000 − Pre-2000)*

We compute the observed change in the probability of positive revisions across the two time periods.

```{r extra_perm_obs_diff, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
obs_diff_pos <- ces_pos |>
  specify(positive_rev ~ post_2000, success = "TRUE") |>
  calculate(stat = "diff in props", order = c("FALSE", "TRUE")) |>
  dplyr::pull(stat)

obs_diff_pos
```

#### 🔹 Null Distribution via Permutation  
*(Assuming no relationship between revisions and time period)*

We randomly shuffle the **post-2000** labels to simulate what differences would look like **if timing didn’t matter**.

```{r extra_perm_null_dist, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
null_dist_pos <- ces_pos |>
  specify(positive_rev ~ post_2000, success = "TRUE") |>
  hypothesize(null = "independence") |>
  generate(reps = 2000, type = "permute") |>
  calculate(stat = "diff in props", order = c("FALSE", "TRUE"))
```

#### 🔹 Two-Sided p-Value  
How extreme is our **observed difference in positive revisions** if **timing had no effect**?

```{r extra_perm_pvalue, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
p_value_pos <- null_dist_pos |>
  get_p_value(obs_stat = obs_diff_pos, direction = "two-sided")

p_value_pos
```

### 🔍 Interpretation of Computational Results

**📌 What do the bootstrap and permutation tests tell us?**

- **Mean Revision (Bootstrap CI):**  
  The 95% bootstrap confidence interval for the mean revision (see `boot_ci_mean`) is very narrow and centered near  
  `r round(obs_mean_rev, 1)` thousand jobs.  
  ➜ This indicates revisions are tiny relative to overall employment — not evidence of major bias.

- **Median Revision (Bootstrap CI):**  
  The bootstrap CI for the median (see `boot_ci_median`) is also close to zero.  
  ➜ Typical revisions are modest and balanced, with extreme outliers not driving the results.

- **Probability of Positive Revisions (Permutation Test):**  
  The permutation test comparing pre- vs post-2000 positive revision rates (`obs_diff_pos` vs. `p_value_pos`) shows no strong statistical shift.  
  ➜ Any differences are easily explained by normal sampling variation, not structural “rigging.”

---

### 🧠 Overall Conclusion from Computational Inference

✔ Revisions remain **small** relative to national employment levels  
✔ Centered around **zero** in both mean and median  
✔ Slight differences over time are **statistically weak**  
✔ Variability aligns with **economic volatility**, not political manipulation

➡ These computational approaches strongly reinforce the findings from earlier statistical tests and fact-checks.